---
layout: default
title: Normalization and Dimensional Reduction
---

<!-- stuff to make Rmarkdown do what we want:  -->
```{r setup, include=FALSE}
suppressMessages(library(rmarkdown))
knitr::opts_chunk$set(echo = TRUE,
  fig.width=9,
  fig.height=6,
  cache=FALSE, # only use TRUE if you're really sure. Also check xfun::cache_rds
  collapse=TRUE,  # put misc. outputs from one chunk in one block
  tidy=TRUE, # tidies up the R code
  tidy.opts=list(arrow=TRUE, 
                 indent=2,
                 width.cutoff = 60))
```

<!-- load complete state from previous lesson -->
```{r loadsession, include=FALSE}
source('../libs.R')
load("../10-hemo/session.rda")
data(refdata_cellranger_GRCh38_3.0.0)
```

# Normalization and Dimensional Reduction

## Normalization

```{r normalize}

srat <- NormalizeData(srat, normalization.method = "LogNormalize")

srat <- ScaleData(srat, features = rownames(srat), verbose = FALSE)

srat <- FindVariableFeatures(srat)

## then SCTransform it (skipping its verbosity with suppressWarnings)
srat <- suppressWarnings(SCTransform(srat, vst.flavor='v2',
                         vars.to.regres = NULL,
                         verbose = FALSE))
  
## As before, we advise using 'vst.flavor='v2' as an additional argument.

```

## Dimensional Reduction

As before, we will reveal some type information of these cells to you,
to make it a bit more interesting. However, it will be only partial:
wether or not the cells are considered to be tumor or healthy.

```{r dimred}

## read the externally provided type information and add it to the object.
file <- paste0(CFG$data_dir, "/rms-types.rds")
types <- readRDS(file)
srat <- AddMetaData(srat, col.name='type', metadata=types)

## run the PCA
srat <- RunPCA(srat, verbose = TRUE, npcs = 50)

  DimPlot(srat,
    reduction = "pca", pt.size = 1, group.by = "plate_id")  |
  DimPlot(srat,
    reduction = "pca", pt.size = 1, group.by = "patient")  |
  DimPlot(srat,
    reduction = "pca", pt.size = 1, group.by = "type")

```

## Choosing the dimensions

```{r choosing-dims}

ElbowPlot(srat, reduction="pca", ndims=50)
##=> around 30?

DimHeatmap(srat, dims = (1:24), cells=100)

## and the next 24:
DimHeatmap(srat, dims = 24 + (1:24), cells=100)
##=> also around 30

CFG$ndims=30

```

## UMAP

Let's do the UMAP and check cell cycle, percentage mitochondria, hemo's
and stress to see if they are might be confounding.

```{r umap}

data(cc.genes) # load the cell cycle genes (also available from SCutils' genelists)

srat<- CellCycleScoring(object = srat,
       s.features =   cc.genes$s.genes,
       g2m.features = cc.genes$g2m.genes,
       assay='RNA')

stress_genes <- genelists$stress # or lookup.stressgenes()
srat <- AddModuleScore(srat, 
                       features=list(stress=stress_genes), 
                       name='stress', 
                       assay='RNA')

srat <- RunUMAP(srat, dims=1:CFG$ndims )

## show the results:
p_plate <- DimPlot(srat, reduction = "umap", pt.size=0.5,
  group.by='plate_id')

p_patient <- DimPlot(srat, reduction = "umap", pt.size=0.5,
  group.by='patient')

p_type <- DimPlot(srat, reduction = "umap",pt.size=0.5,
  group.by='type')

p_phase <- DimPlot(srat, reduction = "umap", pt.size=0.5,
  group.by='Phase')

( p_plate | p_patient ) / ( p_type | p_phase ) 

## and also the continue variables:
## First need colors:
CFG$gradient.colors = viridis::viridis(101, direction= -1)

FeaturePlot(srat, pt.size=0.5, 
  feature=c('nCount_RNA', 'percent_mito',
  'pct_hemo', 'stress1'), order=TRUE,
  cols=CFG$gradient.colors)
 
```

<!-- lastly, save the complete sesssion for the next time -->
```{r savesession, include=FALSE} 
## save complete state for next lesson
## save complete state for next lesson
## save.image("session.rda")
if(Sys.getenv("nodump")!="TRUE")
    save(file="session.rda", list=c("srat", ls(pat="col|CFG")))
```
