---
layout: default
title: Clustering
---

<!-- stuff to make Rmarkdown do what we want:  -->
```{r setup, include=FALSE}
suppressMessages(library(rmarkdown))
knitr::opts_chunk$set(echo = TRUE,
  fig.width=9,
  fig.height=6,
  cache=FALSE, # only use TRUE if you're really sure. Also check xfun::cache_rds
  collapse=TRUE,  # put misc. outputs from one chunk in one block
  tidy=TRUE, # tidies up the R code
  tidy.opts=list(arrow=TRUE, 
                 indent=2,
                 width.cutoff = 60))
```

<!-- load complete state from previous lesson -->

```{r loadsession, include=FALSE} 
source('../libs.R')
load("../12-confound/session.rda")
data(refdata_cellranger_GRCh38_3.0.0)
```

```{r pretend-continuity, include=FALSE} 
### these plots were made in the previous excercise but are too large to 
### include in the session.rda. Therefore, recreate them here.
p_plate <- DimPlot(srat, reduction = "umap", pt.size=0.5,
  group.by='plate_id')

p_patient <- DimPlot(srat, reduction = "umap", pt.size=0.5,
  group.by='patient')

p_type <- DimPlot(srat, reduction = "umap",pt.size=0.5,
  group.by='type')

p_phase <- DimPlot(srat, reduction = "umap", pt.size=0.5,
  group.by='Phase')
### don't plot them!
```

<!-- defaultassay is  still SCT -->

# Clustering

The next step is clustering, i.e. finding distinct groups of cells that
hopefully make sense biologically. Note that clustering is distinct from
creating UMAPs. Often (but not always), the groups of cells 
visible in the UMAP correspond tot he clusters found by `FindClusters`.

Clustering algorithms (of which there are very many) need a parameter
(often called the *resolution*) that controls how many clusters it will
produce. Let's try a few.

```{r clustering}

srat <- FindNeighbors(srat, dims = 1:CFG$ndims)

srat <- FindClusters(srat, resolution = 0.4, algorithm = 1)

srat <- FindClusters(srat, resolution = 0.8, algorithm = 1)

srat <- FindClusters(srat, resolution = 1.2, algorithm = 1)

columns <- grep("SCT_snn_res", names(srat@meta.data), value = TRUE)

## show numbers of cells per cluster (top row shows the
## cluster ids)
for (col in columns) {
  cat("\n====\n", col, ":")
  print(table(srat@meta.data[[col]]))
}

```

<!-- note: you can also pass an array of resolutions -->

## Results

Let's show the results side-by-side to get a feel for it.

```{r choosing-clusresol}

## let's plot the different clusterings side by side
## to see the differeces:

p_clusbyreso <- list() # set up the list to hold the intermediate plots

for (reso in c("0.4", "0.8", "1.2")) {

  clus.column <- paste0("SCT_snn_res.", reso)
  clusters <- sort(unique(srat@meta.data[[clus.column]]))

  ## switching color scheme to make it look different:
  clus.colors <- Polychrome::alphabet.colors(length(clusters))
  ## clus.colors <-
  ## Polychrome::kelly.colors(length(clusters))
  names(clus.colors) <- as.character(clusters)

  p <- DimPlot(srat, pt.size = 1, group.by = clus.column, cols = clus.colors) +
    labs(title = reso)
  p_clusbyreso[[reso]] <- p
}

## use patchwork to plot them side-by-side:
p_clusbyreso[[1]] |  p_clusbyreso[[2]] | p_clusbyreso[[3]] 
## => 0.8 is fine?
```

## Choose the resolution

Based on the above steps it would seem that 0.8 is a reasonable choice,
let's show that:

```{r  }
## set the identies to the clusters coming from the 0.8 resolution:

  clus.column <- paste0("SCT_snn_res.", 0.8)
  Idents(srat) <- clus.column
  
  clusters <- sort(unique(srat@meta.data[[clus.column]]))
  ## switching color scheme to make it look different:
  cluster.colors <- Polychrome::alphabet.colors(length(clusters))
  names(cluster.colors) <- as.character(clusters)

## and show result with the other umaps to get your bearings:
  p_clus <- DimPlot(srat, reduction = "umap", pt.size=0.5, 
     group.by=clus.column, cols=cluster.colors)

( p_plate | p_patient ) / ( p_type | p_clus )

rm(p_clusbyreso)
gc()

```

## Using clustree

Although the best clustering has always to be informed by the biology,
we often need a first guess.  As with the number principal components
previously, there are several rules of thumb to choose the 'best' resolution
for the clustering.

The `clustree` library may be able to help decide this.  It depicts the
'splitting' of clusters as the resolution increases, with
interconnections where cells jump between the main clusters.  A
reasonable resolution is one were the clusters stop splitting up and
there is little 'jumping around' for a few resolutions steps.

For this it is useful to create a larger range of resolutions. 
We can then plot them using the `clustree` function. 

It turns out that `FindClusters` can also accept a vector of
resolutions (this is undocumented).  Let's use that this time, for
brevity.

```{r clustree}

srat <- FindClusters(srat, 
                     resolution = seq(0.4, 2.0, 0.2), 
                     algorithm = 1)

clustree::clustree(srat, prefix = "SCT_snn_res.", node_size=2, edge_width=0.5)

## based on this I think it's still good to choose 0.8
clus.column <- paste0("SCT_snn_res.", 0.8)
Idents(srat) <- clus.column

```

**Exercise 1:** Given these splits in the clustree diagram, see if you can find meaningful gene expression differences that explain the splits.

<!-- lastly, save the complete sesssion for the next time -->
```{r savesession, include=FALSE} 
if(Sys.getenv("nodump")!="TRUE")
    save(file="session.rda", list=c("srat", ls(pat="col|CFG")))
```
