---
layout: default
title: Cell typing with SingleR
---

<!-- stuff to make Rmarkdown do what we want:  -->
```{r setup, include=FALSE}
suppressMessages(library(rmarkdown))
knitr::opts_chunk$set(echo = TRUE,
  fig.width=9,
  fig.height=6,
  cache=FALSE, # only use TRUE if you're really sure. Also check xfun::cache_rds
  collapse=TRUE,  # put misc. outputs from one chunk in one block
  tidy=TRUE, # tidies up the R code
  tidy.opts=list(arrow=TRUE, 
                 indent=2,
                 width.cutoff = 60))
```

<!-- load complete state from previous lesson -->
```{r loadsession, include=FALSE} 
source('../libs.R')
load("../14-enrich/session.rda")
data(refdata_cellranger_GRCh38_3.0.0)
```

<!--  make it look like we had the p_plate etc. plots from previous session.
Including the plots rda file is too big so just remake them but don't show them
-->

```{r pretend-continuity, include=FALSE} 

## make it look like we had the p_plate etc. plots from previous session.
## Including the plots rda file is too big so just remake them but don't show them

clus.column <- paste0("SCT_snn_res.", 0.8) ##

Idents(srat) <- clus.column
  
clusters <- sort(unique(srat@meta.data[[clus.column]]))
## switching color scheme to make it look different:
cluster.colors <- Polychrome::alphabet.colors(length(clusters))
names(cluster.colors) <- as.character(clusters)

p_clus <- DimPlot(srat, reduction = "umap", pt.size=0.5, 
  group.by=clus.column, cols=cluster.colors)

p_plate <- DimPlot(srat, reduction = "umap", pt.size=0.5,
  group.by='plate_id')

p_patient <- DimPlot(srat, reduction = "umap", pt.size=0.5,
  group.by='patient')

p_type <- DimPlot(srat, reduction = "umap",pt.size=0.5,
  group.by='type')

## NOT SHOWING IT, this is just so we have it available later on

```

# Cell typing using SingleR

We already have some idea of what each cluster is, we will now subject it
to [SingleR](https://doi.org/10.1038/s41590-018-0276-y)'s automatic cell
type identification procedure.

The function `plotScoreHeatmap` shows per celltype (rows) and per
cluster of cells (columns) the similarity of a cluster to each of the
celltypes in the reference.  The presence of strong 'blocks' allows you
to assign types per cluster.

```{R run-singler}

## (down)load the reference data:
hpca <- HumanPrimaryCellAtlasData()
## file <- paste0(CFG$data_dir, '/hpca.rds') # cached verssion for speed
## hpca <- readRDS(file=file)

## check how many (and/or which) types:
length( unique(hpca$label.main) )
length( unique(hpca$label.fine) )

## following takes 2 min or so:
singler_hpca <- SingleR(test = GetAssayData(srat, assay = "RNA", slot = "data"),
   ref = hpca, labels = hpca$label.main)

## plot the heatmap
SingleR::plotScoreHeatmap(singler_hpca,
                          show.labels = TRUE, max.labels=100,
                          show.pruned = FALSE,
                          order.by="clusters",
                          clusters = srat@meta.data[[clus.column]], 
                          annotation_colors = list(Clusters=cluster.colors)
                          )
```

## Type Overview

The heatmap shows the cluster ID as the colored strip on top, and the
'best guess' celltypes per cell (not per cluster) as the second strip.
You can find these cell types in the `first.labels` element of the
`singler_hpca` object, and they are in the same order as the cells in
the Seurat object. It's useful add this as meta data. And while we're at
it, let's also do a quick overview of the numbers and create UMAPs of
our newly found celltypes.

```{r singler-overview }

srat@meta.data$singler <- singler_hpca$first.labels

cbind(table(singler_hpca$first.labels)) # cbind to make a nice table

## Plot this:

typenames  <- unique( srat@meta.data$singler  )
singler.colors <- Polychrome::alphabet.colors(length(typenames))
names(singler.colors) <- typenames

p_singler <- DimPlot(srat, reduction = "umap", pt.size=0.5, 
  group.by='singler', cols=singler.colors)

## show it:

p_singler

```

**Exercise 1:** Based on this, what is the most abundant celltype?

**Exercise 2:** Do you think all the types shown were indeed present in this sample?

## Type Assignment

If clusters are sufficiently homogeneous, we often assign the types
based on complete clusters rather than invididual cells like we did
before in the previous UMAP.

Let's call the 'confused clusters' Embryonal Stem Cells (ESC), suffixed
with their cluster-of-origin for clarity.  **Note**: in your R session
the clusters may have been assigned slightly differently, so
double-check the type assignments, the cluster IDs may differ from the
ones shown in the code below.

```{r singler-type-assignment}
## show previous plot to get our bearings:
( p_plate | p_patient ) / ( p_type | p_clus )

## do the asssignment, abbreviating them a bit for convenience 
## and, for the 'suspect' ones, also include the cluster of origin
  
  type2 <- rep("unknown", nrow(srat@meta.data))
  names(type2) <- rownames(srat@meta.data)
  
  type2[ WhichCells(srat, idents = "0"  )  ] <- 'T'
  type2[ WhichCells(srat, idents = "1"  )  ] <- 'ESC-1'
  type2[ WhichCells(srat, idents = "2"  )  ] <- 'ESC-2'
  type2[ WhichCells(srat, idents = "3"  )  ] <- 'ESC-3'
  type2[ WhichCells(srat, idents = "4"  )  ] <- 'NK'
  type2[ WhichCells(srat, idents = "5"  )  ] <- 'ESC-5'
  type2[ WhichCells(srat, idents = "6"  )  ] <- 'Mono'
  type2[ WhichCells(srat, idents = "7"  )  ] <- 'ESC-7'
  type2[ WhichCells(srat, idents = "8"  )  ] <- 'ESC-8'
  type2[ WhichCells(srat, idents = "9"  )  ] <- 'B'
  type2[ WhichCells(srat, idents = "10" )  ] <- 'B'
  type2[ WhichCells(srat, idents = "11" )  ] <- 'Chondro'

srat <- AddMetaData(srat, col.name='type2', metadata=type2)

typenames <- unique(type2)
type2.colors <- Polychrome::alphabet.colors(length(typenames))
names(type2.colors) <- typenames

p_type2 <- DimPlot(srat, reduction = "umap", pt.size=0.5, 
  group.by='type2', cols=type2.colors)

## show the final result, this time together with the other characteristics
## just because we can :-)
( p_patient | p_clus ) / ( p_type | p_type2 )

```

<!-- In the previous run, 'Chondro' was lumped together under the term
  'Mesenchymal'. Do that during the evaluation.
  -->

**Exercise 3:** Do the exact same celltyping, now using the
`hpca$label.fine` instead of `hpca$label.main`

<!-- lastly, save the complete sesssion for the next time -->
```{r savesession, include=FALSE} 
## save complete state for next lesson
## save.image("session.rda")
if(Sys.getenv("nodump")!="TRUE")
    save(file="session.rda", list=c("srat", ls(pat="col|CFG")))
```
