---
layout: default
title: Enrichment analysis
---

<!-- stuff to make Rmarkdown do what we want:  -->
```{r setup, include=FALSE}
suppressMessages(library(rmarkdown))
knitr::opts_chunk$set(echo = TRUE,
  fig.width=9,
  fig.height=6,
  cache=FALSE, # only use TRUE if you're really sure. Also check xfun::cache_rds
  collapse=TRUE,  # put misc. outputs from one chunk in one block
  tidy=TRUE, # tidies up the R code
  tidy.opts=list(arrow=TRUE, 
                 indent=2,
                 width.cutoff = 60))
```

<!-- load complete state from previous lesson -->
```{r loadsession, include=FALSE} 
source('../libs.R')
load("../13-clust/session.rda")
data(refdata_cellranger_GRCh38_3.0.0)
```

# Enrichment analysis

We will now do an enrichment analysis to find out what might be
characteristic for our clusters. This hopefully will give insight into
the role or activity of the cells in the clusters. This analysis will
look for over-representation of 'annotation terms' in the genes that
significantly up in a cluster, relative to the rest of the cells.

First we have to find those characterisic genes per cluster. This is
done based on Differential Expression analysis. In the next step these
genes are analyzed.  There are many ways of doing this, and many
catalogues of 'annotation terms' to do it.  For simplicity, we will only
use the hypergeometric test and [Gene
Ontology](http://geneontology.org/) *Biological Process* terms here.

## Find differentially expressed genes

```{r de-genes}

## first step: find the differentially expressed genes
## this takes ~ 2min
de_genes <- FindAllMarkers(srat, assay='RNA', 
                           only.pos = TRUE,
                           min.pct = 0.1,
                           logfc.threshold = log(1.5))

## If you want to save them for analysis with e.g. http://geneontology.org/,
## you can do something like this:
## 
##   file <- paste0(CFG$output_dir, "/rms_de_genes.txt")
##   write.table(de_genes, file=file, sep = "\t", 
##       quote = FALSE, row.names = T, col.names = NA, na = "")

## The resulting data.frame is too unwieldy to work with, 
## the top 10 genes per cluster are enough:

de_genes %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC) -> top10

## (after this incantatation, top10 contains the slimmed-down list)

## show the contents, maybe you already notice interesting genes:
head(top10)

## show it visually:
DoHeatmap(subset(srat, downsample = 50),
          features = top10$gene,
          group.colors=cluster.colors,
          assay='RNA',
          slot='scale.data'
          ) + theme(axis.text.y=element_text(size=6))

```

## Perform enrichment analysis 

```{r enrichment}

## compareClusters needs a list() of clusters, not the
## grouped_df returned by dplyr::top_n. Convert the top10 to a list:

df  = as.data.frame(top10)
clusters = split(df$gene, df$cluster)

## show contents:
show(clusters)

## compareCluster crashes on gene symbols for some reason,
## so let's use EnsEMBL ids. The SCutils data() statement
## also attaches translation table 'gene_name2ensembl'
## which can be used to do the conversion

## see some contents:
show(sample(gene_name2ensembl, size=10))

## see how it works, e.g. on one of the gene lists:
show(genelists$hemo)
show(gene_name2ensembl[ genelists$hemo ] )

## create a little function that does the conversion for us.
## We omit the original names for readability, and also ignore the NA's 
lookup_ensg_id <- function(id){na.omit(unname(gene_name2ensembl[id]))}

## call it:
clusters <- lapply(clusters, lookup_ensg_id)

## show result:
show(clusters)

## Lastly, run it. This takes ~ 2 min
enrichment <- compareCluster(clusters,
      fun= "enrichGO",
      OrgDb = "org.Hs.eg.db",
      keyType = "ENSEMBL", 
      ont = "BP", minGSSize = 3,
      pAdjustMethod = "BH", 
      pvalueCutoff = 0.01, 
      qvalueCutoff = 0.05,
      readable = TRUE)

## so save it:
file <- paste0(CFG$output_dir, '/enrichment.rds')
saveRDS(file=file, enrichment)

## show results:
clusterProfiler::dotplot(enrichment,
   showCategory = 4, 
   title = "GO biological process",
   font = 5)
## output can be simplified with simplify and dropGO, see the clusterProfiler vignette.

```

**Exercise 1:** Based on this, what would the cell types be?


<!-- lastly, save the complete sesssion for the next time -->
```{r savesession, include=FALSE} 
## save complete state for next lesson
if(Sys.getenv("nodump")!="TRUE")
    save(file="session.rda", list=c("srat", ls(pat="col|CFG")))
```
