---
layout: default
title: Removing confounders
---

<!-- stuff to make Rmarkdown do what we want:  -->
```{r setup, include=FALSE}
suppressMessages(library(rmarkdown))
knitr::opts_chunk$set(echo = TRUE,
  fig.width=9,
  fig.height=6,
  cache=FALSE, # only use TRUE if you're really sure. Also check xfun::cache_rds
  collapse=TRUE,  # put misc. outputs from one chunk in one block
  tidy=TRUE, # tidies up the R code
  tidy.opts=list(arrow=TRUE, 
                 indent=2,
                 width.cutoff = 60))
```

<!-- load complete state from previous lesson -->

```{r loadsession, include=FALSE} 
source('../libs.R')
load("../11-normdimred/session.rda")
data(refdata_cellranger_GRCh38_3.0.0)
```

# Removing confounders

You can remove the usual confounders (see the intestinal organoid
lessons), but this time that is not enough.  Within the main clusters,
there are cell cycle-induced sub-clusters that persist.

The reason is that although the canonical cell cycle genes are gone,
there are more cell cycle genes than the main one and their variability
can still drive the UMAP differences.  We can identify these
extra genes by their correlation with the cell cycle module scores. The
`SCutils` package contains functions to do this, in two steps. The
function `metadataCorrelations` finds, for each gene, the correlation
(over all cells) with  cell meta-data. This can be any numerical
metadata, but in the current case we would use `S.Score` and `G2M.Score`
metadata which were calculated by `Seurat::CellcycleScoring()` in one of
the previous steps.

Next, the function `SCutils::derivedCellcycleGenes()` visualizes this
and returns lists of genes that strongly correlate with `S.Score` and/or
`G2M.Score` but not both[^explain]. See `?metadataCorrelations` and
`?derivedCellcycleGenes` and their source code for details.

Genes in these lists can subsequently be added to the set
of genes that are to be removed from the variable genes, thereby
hopefully achieving a better deconfounding.

[^explain]: We explicity keep the genes that are correlating with both S and G2M phase simultaneously, because they might be 'master switches' for the cellcycle.

```{r deconfounding}

Scor <- metadataCorrelations(srat, 'S.Score')

G2Mcor <- metadataCorrelations(srat, 'G2M.Score')

additional <- derivedCellcycleGenes(Scor=Scor,
                                    Sgenes=genelists$s.genes,
                                    G2Mcor=G2Mcor,
                                    G2Mgenes=genelists$g2m.genes,
                                    plot=TRUE)

## see what's in the lists:
show(additional)

## also get the other deconfounders:
stress_genes <- genelists$stress # OR: SCutils::lookup.stressgenes()
hb_genes <- genelists$hemo # as before

## combine these sets (also including stress_genes and hemoglobin genes)
remove <- unique(c(cc.genes$s.genes, cc.genes$g2m.genes, 
  additional$S.derived, additional$G2M.derived, 
  stress_genes, hb_genes))

## check how many we will loose now:
length(intersect(VariableFeatures(srat), remove))

```

## Redo the Dimensional Reduction

We can now throw out the confounders. 
After this, we have to do the complete dimensional reduction over again!

```{r remove-vargenes}

## do the removal, for both the 'RNA' (i.e. LogNorm) and 'SCT' assays:
VariableFeatures(srat, assay='RNA') <-
   setdiff( VariableFeatures(srat, assay='RNA'), remove)

VariableFeatures(srat, assay='SCT') <- 
   setdiff( VariableFeatures(srat, assay='SCT'), remove)

```

**Exercise 1:** How many variables features are removed / remain ? Can you explain the difference in numbers?

<!-- 
length(VariableFeatures(srat, assay='RNA'))
length(VariableFeatures(srat, assay='SCT'))
-->

```{r rerun-dimred}
srat <- RunPCA(srat, verbose = TRUE, npcs = 50)

srat <- RunUMAP(srat, dims=1:CFG$ndims )

## show the results:
p_plate <- DimPlot(srat, reduction = "umap", pt.size=0.5,
  group.by='plate_id')

p_patient <- DimPlot(srat, reduction = "umap", pt.size=0.5,
  group.by='patient')

p_type <- DimPlot(srat, reduction = "umap",pt.size=0.5,
  group.by='type')

p_phase <- DimPlot(srat, reduction = "umap", pt.size=0.5,
  group.by='Phase')

## use patchwork again to get a 2 x 2 plot
( p_plate | p_patient ) / ( p_type | p_phase )

## and also the continuous variables:
FeaturePlot(srat, pt.size=0.5, 
  feature=c('nCount_RNA', 'percent_mito',
  'pct_hemo', 'stress1'), order=TRUE,
  cols=CFG$gradient.colors)

```

**Exercise 2:** Make side-by-side UMAP plots before and after deconfounding, and compare to see if we indeed got rid of confounding.

**Exercise 3:** Is sex a confounder for these samples? 
You can use `genelists$male` and/or `genelists$female` for this. 
Same question for the ribosomal protein genes, using `genelists$ribo`
(or any of the equivalent `lookup.\*` functions).

<!-- lastly, save the complete sesssion for the next time -->
```{r savesession, include=FALSE} 
## save complete state for next lesson
## save.image("session.rda")
if(Sys.getenv("nodump")!="TRUE")
    save(file="session.rda", list=c("srat", ls(pat="col|CFG")))
```
