---
layout: default
title: Clustering
---

<!-- stuff to make Rmarkdown do what we want:  -->
```{r setup, include=FALSE}
suppressMessages(library(rmarkdown))
knitr::opts_chunk$set(echo = TRUE,
  fig.width=9,
  fig.height=6,
  cache=FALSE, # only use TRUE if you're really sure. Also check xfun::cache_rds
  collapse=TRUE,  # put misc. outputs from one chunk in one block
  tidy=TRUE, # tidies up the R code
  tidy.opts=list(arrow=TRUE, 
                 indent=2,
                 width.cutoff = 60))
```

<!-- load complete state from previous lesson -->
```{r loadsession, include=FALSE} 
source('../libs.R')
load("../06-markergenes/session.rda")
gc()

```

# Clustering

The data as presented in the UMAP's shows clusters visually, but we
cannot really work with them. We need to have a actual assignment of
cells to clusters in order to e.g. find genes that are differentially
expressed between clusters. Keep in mind that some of the clustering
behaviour is due to confounding (as we already saw), whereas the rest is
actual biology.

There are many clustering algorithms, with many parameters to
tweak. Because single-cell data sets are huge, speed is required which
is achieved by first computing the network of approximate nearest
neighbors ('ANN') using the `FindNeighbors` function.  `FindClusters`
then uses this network to establish the clustering. `FindClusters`
offers 4 different methods. The default method (Louvain clustering,
algorithm=1) is fine and runs everywhere. 

In the ideal case, clusters are 'obvious': each cluster is a little
island, and all the islands are well separated. But frequently things
are less clear-cut, with untidy continents and peninsulas connected by
sprawling archipelagos.  The algorithm has to make choices on where to
divide bigger groups into separate clusters.  This of course depends on
the data, but also on a parameter we have to supply, namely the
resolution. A low resolution will give fewer but larger clusters; 
a high resolution yields more but smaller clusters. Again, you have
to play around a bit. 

Seurat stores the cluster assignments (simple integers) in the `meta.data`
slot, each under a name constructed from the `DefaultAssay` and the resolution.
This allows you to easily compare clusterings. In addition, the most recent
clustering is stored in the `seurat_clusters` column. 

```{r clustering}

srat <- FindNeighbors(srat,  dims = 1:CFG$ndims)
                              
srat <- FindClusters(srat, resolution = 0.4,
                              algorithm = 1)

srat <- FindClusters(srat, resolution = 0.8,
                              algorithm = 1)

srat <- FindClusters(srat, resolution = 1.2,
                              algorithm = 1)

## show the columns that were added to the metadata:
head(srat@meta.data)

## or
columns <- grep("SCT_snn_res", names(srat@meta.data), value=TRUE)
head(srat@meta.data[ , columns])

## show numbers of cells per cluster (top row shows the cluster ids)
for (col in columns) {
    cat("\n====\n", col, ":")
    print(table(srat@meta.data[[col]]))
}

```

Let's show the clusters side-by-side with the cell types we gave you. 

<!--  The assignment to list() is partly because it wouldn't plot ... -->

```{r plot-clusters}

## assign plot to a variable for convenience:
p_bytype <- DimPlot(srat, pt.size=1, group.by='type', cols=type.colors) 

## for easy plotting, assign the different resolution plots to a list() 
p_clusbyreso <- list()

for (reso in c("0.4",  "0.8", "1.2" )) { 
 
 column <- paste0("SCT_snn_res.", reso)
 clusters <-sort(unique(srat@meta.data[[column]]))
 
 ## switching color scheme to make it look different:
 clus.colors <- Polychrome::alphabet.colors(length(clusters))
 names(clus.colors) <- as.character(clusters)

 p <- DimPlot(srat, pt.size=1, group.by=column, cols=clus.colors ) +  labs(title=reso)
 p_clusbyreso[[reso]]<- p
}

## plot them side by side:

p_bytype | p_clusbyreso[[ "0.4" ]] 

p_bytype | p_clusbyreso[[ "0.8" ]] 
 
p_bytype | p_clusbyreso[[ "1.2" ]] 

```

```{r rm_p_clusbyreso, include=FALSE} 

rm(p_clusbyreso)
gc()

```

Most clusters correspond to identifiable groups in the UMAP.
Based on this, let's fix the resolution at 1.2 and proceed.

<!-- lastly, save the complete sesssion for the next time -->
```{r savesession, include=FALSE} 
## save complete state for next lesson
## save.image("session.rda")
if(Sys.getenv("nodump")!="TRUE")
    save(file="session.rda", list=c("srat", ls(pat="col|CFG")))
```
